Bayesian non-parametric inference for multivariate peaks-over-threshold models

### 2023-07-07 results.tex
prep for submission

Our hyperprior parameters are set as follows: for the gamma-centered models (PG--G, PRG--G), 
    the shape parameter for the centering 
    distribution~$\xi_{\ell}\sim \mathcal{G}\left(1,1\right)$, and
    rate parameter~$\tau_{\ell}\sim\mathcal{G}\left(2,2\right)$.
    For the log-normal centered models (PG--LN, PRG--LN), the centering distribution's
    log-mean $\bm{\mu}\sim\mathcal{N}_d\left(0,\bm{I}_d\right)$, and covariance 
    matrix~$\Sigma\sim\mathcal{IW}\left(d + 10, (d+10)\bm{I}_{d}\right)$.  These values are 
    intended such that draws from the prior for $\Sigma$ will weakly tend towards the 
    identity matrix.
    For models learning rate parameters $\beta_{j\ell}$ (PG--G, PG--LN), the centering
    distribution's shape parameter $\zeta\sim\mathcal{G}\left(1,1\right)$ and rate parameter
    $\sigma\sim\mathcal{G}\left(2,2\right)$.  The choice of the $\mathcal{G}(2,2)$ for rate
    parameters places no mass near 0, in order to draw estimates for the value away from 
    0 for numerical stability.

\subsection{Simulation Study\label{subsec:simulated}}
To evaluate our proposed approach for angular measure estimation 
    we consider simulated datasets on $\mathbb{S}_{\infty}^{d-1}$,
    for values of $d$ ranging between 2 and 32.  We generated each
    dataset as a mixture of multivariate log-normal distributions,
    projected onto $\mathbb{S}_{\infty}^{d-1}$. The generation 
    procedure is detailed in Algorithm~\ref{algo:simulated}.  We consider two gamma-centered
    and two log-normal centered DP mixture models, with and without restrictions in each
    case. We also consider a Dirichlet distribution and 
    a pairwise betas model \citep{COOLEY2010}, fitted using the \verb|BMAmevt| 
    package in R~\citep{BMAmevt}. For the DP mixture
    models, the data are projected onto $\mathbb{S}_{10}^{d-1}$. For the other
    two models they are projected on $\mathbb{S}_1^{d-1}$.

@@

Figure~\ref{fig:simpples} shows the average rise over baseline in energy score as
    calculated on $\mathbb{S}_{\infty}^{d-1}$ using the kernel metric described
    in Proposition~\ref{prop:rot}, for models trained on simulated data.  After training
    a model, a posterior predictive dataset is generated, and the energy score is
    calculated as a Monte Carlo approximation of Equation~\ref{eq:es}.
    The \emph{baseline} value is the energy score of a new dataset from the same
    generating distribution as the training dataset, evaluated against the training 
    dataset.  
%    We considered a total of four variations of the Dirichlet-process mixture
%    of projected gammas model: whether to restrict $\beta_{j\ell} := 1$ as the projected
%    \emph{restricted} gamma (DP-PRG) akin to a projected Dirichlet, or allow them to be 
%    learned with a gamma prior (DP-PG); and for the shape parameters whether to use a 
%    gamma centering distribution (-G) or a log-normal centering distribution (-LN).
%    For comparison we also included two other distributions on $\mathbb{S}_{1}^{d-1}$:
%    The Dirichlet distribution (with gamma priors), and the pairwise betas distribution
%    developed in~\cite{COOLEY2010}.  
    % Note that the pairwise betas model respects
    % the moment restriction generally assumed.\makenote{This phrasing needs to be changed.
    % The point I am trying to make is that because the moment-restriction assumption is 
    % discarded, a moment-restricted model is inadequate, as is borne out in the
    % simulation study.} \bruno{\bf I would avoid getting into the details of what the
    % pairwise beta model does. More important is to say why we chose it, and how available 
    % it is}
    \makenote{It was anticipated that the log-normal centering distribution would place
    more mass of the  DP into regions of the support with active clusters, which would 
    generate new clusters more readily.  For the simulated data,} 
    we observe that the projected gamma models dominate, and within that class
    of model, the choice of centering distribution is immaterial.  For the 
    projected \emph{restricted} gamma models, the choice of a multivariate log-normal 
    centering distribution appears to \makenote{somewhat} negatively affect model 
    fidelity.  \makenote{Additionally, the log-normal models with the additional computational cost
    of multivariate normal evaluation and parallel tempering, take approximately six times
    longer to sample relative to the gamma models.}
    % \bruno{The comparison needs some information about the computing times.}

@@

\begin{figure}[tb]
    \begin{center}
    \caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
    \begin{subfigure}[b]{0.48\textwidth}
        \centering 
        \includegraphics[height=2.5in]{./images/erai_grid}
    \end{subfigure}
    %
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \includegraphics[height=2.5in]{./images/era5_grid}
    \end{subfigure}
    \end{center}
\end{figure}

@@

% Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy 
%     scores for the different models considered. We observe that, as in the case of the simulation 
%     study shown in Figures~\ref{fig:simpples}, the preferred model is the projected restricted 
%     gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
%     Interestingly, this preference for the restricted gamma form is the opposite of the result 
%     obtained in the simulation example.
Table~\ref{tab:dev} shows the values of the estimated energy 
    scores for the different models considered. We observe that, contrary to the results in the
    simulation study in Figure~\ref{fig:simpples}, the preferred model is the projected restricted 
    gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.

@@

We consider an exploration of the pairwise extremal dependence using Monte Carlo estimates of the 
  coefficients in  Equation~\ref{eqn:chi_ij}. For this we use samples obtained from the PRG--G model.
  Figure~\ref{fig:chi_ij} provides a graphical analysis of the results. 
  The coefficients achieve values between $0.286$ and $0.759$ for the ERA-Interim data and 
  between $0.181$ and $0.840$ for the ERA5 data.  The greater range in dependence scores observed 
  with the ERA5 data versus ERA-Interim speaks to the greater granularity of the ERA5 data--distance between locations is a strong contributor to the strength of the 
  pairwise asymptotic dependence. The highest coefficients are $0.759$ for 
  locations 4 and 5 in the ERA-Interim data and
  $0.840$ for locations 1 and 2 in the ERA5 data.  Clearly, pairwise asymptotic
  dependence coefficients tell a limited story, as a particular dependence may include
  more than two locations.   We can, however, glean some information from the patterns that
  emerge in two dimensions.  For the ERA-Interim data, we observe a possible cluster 
  between cells 5-8, indicating a strong dependence among these cells.  Analogously, for
  the ERA5 data, we observe three possible groups of locations.

>>>>

Our hyperprior parameters are set as follows: for the gamma-centered models (PG--G, PRG--G), 
    the shape parameter for the centering 
    distribution~$\xi_{\ell}\sim \mathcal{G}\left(1,1\right)$, and
    rate parameter~$\tau_{\ell}\sim\mathcal{G}\left(2,2\right)$.
    For the log-normal centered models (PG--LN, PRG--LN), the centering distribution's
    log-mean $\bm{\mu}\sim\mathcal{N}_d\left(0,\bm{I}_d\right)$, and covariance 
    matrix~$\Sigma\sim\mathcal{IW}\left(d + 10, (d+10)\bm{I}_{d}\right)$.  These values are 
    intended such that draws from the prior for $\Sigma$ will weakly tend towards the 
    identity matrix.
    For models learning rate parameters $\beta_{j\ell}$ (PG--G, PG--LN), the centering
    distribution's shape parameter $\zeta\sim\mathcal{G}\left(1,1\right)$ and rate parameter
    $\sigma\sim\mathcal{G}\left(2,2\right)$.  The choice of the $\mathcal{G}(2,2)$ for rate
    parameters places little mass near 0, in order to draw estimates for the value away from 
    0 for numerical stability.

\subsection{Simulation Study\label{subsec:simulated}}
To evaluate our proposed approach for angular measure estimation 
    we consider simulated datasets on $\mathbb{S}_{\infty}^{d-1}$,
    for values of $d$ ranging between 2 and 32.  We generated each
    dataset as a mixture of multivariate log-normal distributions,
    projected onto $\mathbb{S}_{\infty}^{d-1}$. The generation 
    procedure is detailed in Algorithm~\ref{algo:simulated}. We produced 
    ten replicates of each configuration. We consider two gamma-centered
    and two log-normal centered DP mixture models, with and without restrictions in each
    case. To perform a comparative analysis we fitted 
   % \makenote{Comparative analysis using posterior predictive samples necessitates
    %alternative models having defined methods of sampling from their respective posterior
    %predictive distributions, for models defined on some normalization of the positive cone.  
    %To that end, we also consider the Dirichlet distribution, as well as} 
    %We also consider a Dirichlet distribution and 
    the pairwise betas model proposed in \cite{COOLEY2010}. We chose this model for comparison
    as it is implemented in the readily available package \verb|BMAmevt| 
    in R~\citep{BMAmevt}. \verb|BMAmevt| provides samples from the predictive
    posterior distribution. These are needed for the calculation of the energy
    scores that are at the basis of our comparison. In addition, \verb|BMAmevt| can 
    be fitted to moderately large multivariate observations. For the DP mixture
    models, the data are projected onto $\mathbb{S}_{10}^{d-1}$. For the other
    two models they are projected on $\mathbb{S}_1^{d-1}$.   We sampled each model for 50,000 
    iterations, dropping the first 40,000 as burn-in, and thinning to keep every 10th iteration
    after.  These settings were intended to provide a consistent sampling strategy that would
    work with every model, even if inefficient for some.

@@

Figure~\ref{fig:simpples} shows the average rise over baseline in energy score as
    calculated on $\mathbb{S}_{\infty}^{d-1}$ using the kernel metric described
    in Proposition~\ref{prop:rot}, for models trained on simulated data.  After training
    a model, a posterior predictive dataset is generated, and the energy score is
    calculated as a Monte Carlo approximation of Equation~\ref{eq:es}.  In our analysis,
    after burn-in and thinning, we had 1,000 replicates from the posterior distribution, and
    generated 10 posterior predictive replicates per iteration.
    The \emph{baseline} value is the energy score of a new dataset from the same
    generating distribution as the training dataset, evaluated against the training 
    dataset.  
%    We considered a total of four variations of the Dirichlet-process mixture
%    of projected gammas model: whether to restrict $\beta_{j\ell} := 1$ as the projected
%    \emph{restricted} gamma (DP-PRG) akin to a projected Dirichlet, or allow them to be 
%    learned with a gamma prior (DP-PG); and for the shape parameters whether to use a 
%    gamma centering distribution (-G) or a log-normal centering distribution (-LN).
%    For comparison we also included two other distributions on $\mathbb{S}_{1}^{d-1}$:
%    The Dirichlet distribution (with gamma priors), and the pairwise betas distribution
%    developed in~\cite{COOLEY2010}.  
    % Note that the pairwise betas model respects
    % the moment restriction generally assumed.\makenote{This phrasing needs to be changed.
    % The point I am trying to make is that because the moment-restriction assumption is 
    % discarded, a moment-restricted model is inadequate, as is borne out in the
    % simulation study.} \bruno{\bf I would avoid getting into the details of what the
    % pairwise beta model does. More important is to say why we chose it, and how available 
    % it is}
   % \makenote{It was anticipated that the log-normal centering distribution would place
    %more mass of the  DP into regions of the support with active clusters, which would 
    %generate new clusters more readily.}  
    For the simulated data, we observe that the projected gamma models dominate the other two
    options considered, regardless of the choice of centering distribution.  The  projected 
    restricted gamma models with a multivariate log-normal centering distribution appears to 
    be dominated by the models based on the alternative centering distributions. Moreover, the
    performance deteriorates with the increase in dimensionality. Additionally, models
    centered around the log-normal distribution incur in the computational cost of 
    multivariate normal evaluation and parallel tempering, taking approximately six times
    longer to sample relative to the gamma models.  We also note that the 
    computational cost  of the pairwise betas model grows combinatorially, with a 
    sample space of dimension $\binom{d}{2} + 1$. In our testing, \verb|BMAmevt| is 
    substantially faster than any of our proposed DP mixture models for low dimensions, 
    however for examples with high number of dimensions, the computational time for 
    \verb|BMAmevt| was comparable or greater than that for DP-PG-G.  We compare computing
    times in our data analysis in Table~\ref{tab:sampletime}.
    % \bruno{The comparison needs some information about the computing times.}

@@

\begin{figure}[tb]
    \caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
    \centering
    \begin{minipage}{0.25\textwidth}
    \centering
    \includegraphics[width=0.99\linewidth]{./images/erai_grid}
    \end{minipage}%
    \begin{minipage}{0.25\textwidth}
    \centering
    \includegraphics[width=0.99\linewidth]{./images/era5_grid}
    \end{minipage}  
    % \caption{Grid cell locations for ERA-Interim (left) and ERA5 (right).\label{fig:gridlocs}}
    % \begin{center}
    % \begin{subfigure}{0.48\textwidth}
    %     \centering 
    %     \includegraphics[height=2.5in]{./images/erai_grid}
    % \end{subfigure}
    % %
    % \begin{subfigure}{0.48\textwidth}
    %     \centering
    %     \includegraphics[height=2.5in]{./images/era5_grid}
    % \end{subfigure}
    % \end{center}
\end{figure}

@@

\begin{table}[htb]
  \centering
  \caption{Time to sample (in minutes) 50,000 iterations for various models\label{tab:sampletime}}
  \include{./tables/table_sampletime}
\end{table}

Table~\ref{tab:dev} shows the values of the estimated energy 
    scores for the different models considered. We observe that, contrary to the results in the
    simulation study in Figure~\ref{fig:simpples}, the preferred model is the projected restricted 
    gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
    Table~\ref{tab:sampletime} shows the computing times needed to fit the different models to the
    two datasets.  We see the effect of dimensionality on the various models; for gamma centered
    models it grows linearly; for the log--normal centered model, it will grow superlinearly as matrix
    inversion becomes the most costly operation.  For \verb|BMAmevt|, its parameter space grows
    combinatorically with the number of dimensions, and thus so does computational complexity and
    sampling time.
    % \bruno{Insert here the table that you sent me in the email and comment the results.}

@@

We consider an exploration of the pairwise extremal dependence using Monte Carlo estimates of the 
  coefficients in  Equation~\ref{eqn:chi_ij}. For this we use samples obtained from the PRG--G model.
  Figure~\ref{fig:chi_ij} provides a graphical analysis of the results. 
  The coefficients achieve values between $0.286$ and $0.759$ for the ERA-Interim data and 
  between $0.181$ and $0.840$ for the ERA5 data.  The greater range in dependence scores observed 
  with the ERA5 data versus ERA-Interim speaks to the greater granularity of the ERA5 data,
  indicating that distance between locations is a strong contributor to the strength of the 
  pairwise asymptotic dependence. The highest coefficients are $0.759$ for 
  locations 4 and 5 in the ERA-Interim data and
  $0.840$ for locations 1 and 2 in the ERA5 data.  Clearly, pairwise asymptotic
  dependence coefficients tell a limited story, as a particular dependence may include
  more than two locations.   We can, however, glean some information from the patterns that
  emerge in two dimensions.  For the ERA-Interim data, we observe a possible cluster 
  between cells 5-8, indicating a strong dependence among these cells.  Analogously, for
  the ERA5 data, we observe three possible groups of locations.
